---
output: github_document
---

<!-- readme.md is generated by readme.rmd, please edit that document instead -->

# theme61

```{r echo = FALSE, message = FALSE, warning = FALSE}
devtools::load_all()
library(theme61)
library(ggplot2)
library(lubridate)
library(data.table)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

This package contains functions and colour palettes that enable you to create e61 Institute themed graphs and know what colours to use in other applications.

### Installing theme61

If you haven't already, install the `devtools` package:

`install.packages::devtools()`

Then install from GitHub:

`devtools::install_github("aaronw22/theme61", dependencies = TRUE, upgrade = "always")`

Then load into R:

`library(theme61)`

### Making the best use of the features

See the [package website](https://e61-institute.github.io/theme61/) to learn how to use the package features.

### Full colour palette

```{r echo = FALSE, message = FALSE, warning = FALSE}
inputs <- CJ(row = seq(1, 12), col = seq(0, 8))
inputs[, value := .I][, value := as.factor(value)]
inputs[, row := factor(row, labels = rev(c("Blue dark", "Blue light", "Teal dark", "Teal light", "Sky light", "Sky dark", "Coral dark", "Coral light", "Orange light", "Orange dark", "Maroon dark", "Maroon light")))]

prep_pal <- function(search) {
  unname(unlist(mget(apropos(search), inherits = TRUE)))
}

col_pal <- lapply(c("e61_bluedark", "e61_bluelight", "e61_tealdark", "e61_teallight", "e61_skylight", "e61_skydark", "e61_coraldark", "e61_corallight", "e61_orangelight", "e61_orangedark", "e61_maroondark", "e61_maroonlight"), prep_pal)
col_pal <- rev(col_pal)
col_pal <- unlist(col_pal)

ggplot(inputs, aes(x = col, y = row, fill = value)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_manual(values = col_pal) +
  scale_x_continuous(n.breaks = 9) +
  theme_bw() +
  theme(legend.position = "none", panel.grid = element_blank(), axis.title = element_blank(), panel.border = element_blank(), axis.ticks = element_blank())
```

