---
title: "Adding on-graph labels"
output: rmarkdown::html_vignette
description: Some example code showing how to put labels on graphs
vignette: >
  %\VignetteIndexEntry{Adding on-graph labels}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
devtools::load_all()
library(ggplot2)
library(ggrepel)
library(data.table)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

## Why labels?

Coloured labels placed directly on the plot are better than graph legends when you are making the final, polished version of a graph for presentations, notes and papers for three key reasons.

Labels make it much easier for the reader to identify what group each colour corresponds to, without having to do the rapid eye twitch from the graph to the legend and back again. Repeat this for every group present on the graph and soon you have a recipe for eyestrain.

Labels make more efficient use of the available space on a graph by removing the need to make room for a legend. Legends create unused white space around the legend element and force the plot to be shorter or narrower than it otherwise would be.

Finally, labels are more aesthetic than legends.

## How-to guide

### Simple cases using `plot_label()`

The simplest way to add plot labels is to use the `plot_label()` function in `theme61`. The following code provides an example. Most of the code is to set up the mock data and create a graph, the only two lines that are important are the `plot_label()` lines.

**If you have more than one label you should use `mplot_label()`**, which is introduced in the next section.

`plot_label()` requires you to provide the text of the label (`label`), the `x` and `y` positioning, the total number of different colours/labels to plot (`n_labs`), and which number element you are currently plotting (`n`). For more information, see the documentation (`?plot_label`).

```{r}
# Mock data
graph_data <- CJ(
  x = 2000:2020,
  source = c("Source A", "Source B")
)

graph_data[source == "Source A", y := seq(0, by = 0.25, length = 21)]
graph_data[source == "Source B", y := seq(0, by = 1, length = 21)]

# Graph
graph <- ggplot(graph_data, aes(x, y, colour = source)) +
  geom_line(linewidth = 1) +
  plot_label(label = "Source A", x = 2017, y = 6, n_labs = 2, n = 1) +
  plot_label(label = "Source B", x = 2017, y = 13, n_labs = 2, n = 2) +
  theme_e61() +
  scale_y_continuous_e61(limits = c(0, 20, 5)) +
  scale_colour_e61(2) +
  labs_e61(
    title = "Simple Case Using plot_label()",
    sources = c("e61 Institute"),
    x = NULL, y = "n"
  )

save_e61(here::here("man/figures/label-fig-1.svg"), graph)
knitr::include_graphics(here::here("man/figures/label-fig-1.svg"))

```

### Adding multiple labels using `mplot_label()`

When you have multiple plot labels, `mplot_label()` allows you to add them in one function. 

When using this function, make sure the wording of the vector of label names in the label argument are exactly the same as the colour/fill aesthetic being plotted, and the order of the label vector matches the order of the colour/fill aesthetic for the ordering of the colours.  

One way to ensure this is by setting the column in the data to be an ordered factor, and calling the label names using `levels(data$col_name)`.

```{r}
# Mock data
graph_data <- CJ(
  x = 2000:2020,
  source = ordered(c("Source A", "Source B"))
)

graph_data[source == "Source A", y := seq(0, by = 0.25, length = 21)]
graph_data[source == "Source B", y := seq(0, by = 1, length = 21)]

# Graph
graph <- ggplot(graph_data, aes(x, y, colour = source)) +
  geom_line(linewidth = 1) +
  mplot_label(
    label = levels(graph_data$source),
    x = c(2017, 2017),
    y = c(6, 13)
  ) +
  theme_e61() +
  scale_y_continuous_e61(limits = c(0, 20, 5)) +
  scale_colour_e61(2) +
  labs_e61(
    title = "Multiple Labels Using mplot_label()",
    sources = c("e61 Institute"),
    x = NULL, y = "n"
  )

save_e61(here::here("man/figures/label-fig-2.svg"), graph)
knitr::include_graphics(here::here("man/figures/label-fig-2.svg"))

```

### Dealing with facets

When you have a faceted graph, using `plot_label()` or `mplot_label()` will not work properly. This is because the annotations will appear in the same position on all the facets and there is no way to prevent this. This is problematic because the labels are unnecessarily repeating themselves, and the correct position of the text on one facet may be different on each facet (e.g. if the height of the columns differs across facets).

There is a way to get around this problem using the `data.frame` approach to labelling the graph.

The `data.frame` approach uses a helper `data.frame` that contains all the information needed to plot the text and then uses `geom_text()` to add the text to the graph. 

In addition to the label text, x and y positions, you also need to include a `facet` column in the `data.frame`. Note that we only provide one of the facets, in this case "A". Notice that we also set this column to be a factor column with **two** levels, "A" and "B". This step is important, as what it does internally is plot the graph label on facet "A", and then when it goes to plot the labels on facet "B" they are missing and hence nothing is plotted.

As an aside, if you want to put one label on the first facet and the second label on another facet, then in the `facet` column, simply allocate the first label to "A" and the second to "B".

```{r}
# Mock data
graph_data <- data.frame(
  facet = c("A", "A", "B", "B"),
  x = 1,
  y = c(10, 15, 10, 15),
  group = c("P", "Q", "P", "Q")
)

graph_labels <- data.frame(
  x = c(0.75, 1.25),
  y = c(12, 17),
  group = c("P", "Q"),
  facet = factor("A", c("A", "B"))
)

# Graph
graph <- ggplot(graph_data, aes(x, y, fill = group)) +
  facet_wrap(~facet) +
  geom_col(position = "dodge") +
  # This geom_text adds the labels
  # The label arg add the text, the colour arg ensures the text colour matches 
  # the lines they are identifying.
  geom_text(data = graph_labels, aes(x, y, label = group, colour = group)) +
  theme_e61() +
  scale_y_continuous_e61(limits = c(0, 20, 5)) +
  scale_colour_e61(2) +
  scale_fill_e61(2) +
  labs_e61(
    title = "Facetted Case Using a Helper data.frame",
    sources = c("e61 Institute"),
    x = NULL, y = "n"
  )

save_e61(here::here("man/figures/label-fig-3.svg"), graph)
knitr::include_graphics(here::here("man/figures/label-fig-3.svg"))

```

### Dealing with many categories

Sometimes you have many categories, far too many to want to plot manually even with the `data.frame` approach. If you are willing to sacrifice the ability to specify the exact location of each label, you could use `geom_text_repel()` to add appropriately coloured text labels for each colour/fill aesthetic. 

```{r}
# Mock data
graph_data <- CJ(
  year = 2000:2020,
  source = paste("Source", LETTERS[1:5])
)
graph_data[, value := sample(5, nrow(graph_data), replace = TRUE)]

# Graph
graph <- ggplot(graph_data, aes(x = year, y = value, fill = source)) +
  geom_area(linewidth = 1) +
                  # We only need one label for each group, ideally on the far 
                  # right of the graph, so we keep the last year only
  geom_text_repel(data = graph_data[year == max(year)],
                  aes(x = year, y = value, label = source, colour = source),
                  # This forces the text to appear further right than 2020
                  xlim = c(2020, NA),
                  # This moves the vertical position of the label downwards
                  position = position_stack(vjust = 0.5),
                  # This hides the line that joins the label to the graph element
                  segment.color = NA,
                  size = 4) +
  theme_e61(fix_left = 3) +
  # Extends the right-side of the graph to create room for the labels
  scale_x_continuous(limits = c(NA, 2026)) +
  scale_y_continuous_e61(limits = c(0, 25, 5)) +
  scale_fill_e61(5) +
  scale_colour_e61(5) +
  labs_e61(
    title = "Use geom_text_repel to Show Many Labels Automatically",
    sources = c("e61 Institute"),
    x = NULL, y = "n"
  )

save_e61(here::here("man/figures/label-fig-4.svg"), graph)
knitr::include_graphics(here::here("man/figures/label-fig-4.svg"))

```
