---
title: "Introduction to theme61"
output: rmarkdown::html_vignette
description: Some example code showing how to use `theme61` with `ggplot2` graphs.
vignette: >
  %\VignetteIndexEntry{Introduction to theme61}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
devtools::load_all()
library(ggplot2)
library(data.table)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`theme61` is an extension to `ggplot2`. It is made up of a custom theme and functions that change the appearance of a standard `ggplot2` graph. Thus, the correct usage of this package is to create your normal `ggplot2` graph, then add `theme61`-specific elements. 

You should read this article and the [Graph standards](graph-standards.html) article to help you understand how to create graphs following the e61 graph standards. 

This article provides some examples to help you figure out what elements you can change. The function documentation provides comprehensive information on the full customisability of the package.

## The colour palette

```{r colour-palette-chart, echo = FALSE}
source("colour-palette.R", print.eval = TRUE)
```

## Correct workflow

When making graphs, you should follow the below workflow. 

Save your graph using `save_e61()`. Do **not** use the Export button in the Plots tab or `ggsave()`.

Start by creating an unthemed graph using the 'normal' stuff: `ggplot(data, aes(x, y)) + geom_blah()`

Then:

- Apply the theme: `theme_e61()`
- Add graph colours: `scale_colour_e61()` or `scale_fill_e61()`
- Format the x and y axes scales: `scale_y_continuous_e61()`, `scale_x_continuous_e61()`
- Add plot labelling: `mplot_label()` or `plot_label()`
- Add titles, axes labels, footnotes and sources: `labs_e61(title = "Title", footnotes = "Footnotes", sources = c("Source 1", "Source 2"))`
- Create a multi-panel (if required): `mpanel_e61()`
- Save the graph: `save_e61()`

The [Graph standards](graph-standards.html) article sets out this workflow in greater detail.

## Graph examples

### Scatter graph

```{r}

# Set up data
mtcars2 <- within(mtcars, {
  vs <- factor(vs, labels = c("V-shaped", "Straight"))
  am <- factor(am, labels = c("Automatic", "Manual"))
  cyl  <- factor(cyl)
  gear <- factor(gear)
})

# Graph
p1 <- ggplot(mtcars2) +
  geom_point(aes(x = wt, y = mpg, colour = gear)) +
  labs_e61(
    title = "Fuel Economy Declines as Weight Increases",
    subtitle = "1973-74",
    sources = "1974 Motor Trend US magazine.",
    x = "Weight (1000 lbs)",
    y = "mpg",
    colour = "Gears"
  )


graph <- p1 +
  theme_e61(adj = -12) + 
  mplot_label(label = c("3", "4", "5"),
              x = c(4, 2.7, 2.5),
              y = c(23, 35, 15)) +
  scale_y_continuous_e61(limits = c(0, 40)) +
  scale_colour_e61(n = 3) +
  add_e61_logo() + 
  square_legend_symbols()

save_e61(here::here("man/figures/fig-1.svg"), graph)

knitr::include_graphics(here::here("man/figures/fig-1.svg"))

```

### Faceted scatter graph

```{r}
graph <- p1 +
  facet_grid(vs ~ am) +
  theme_e61(legend = "bottom", legend_title = TRUE) +
  scale_colour_e61(n = 3) +
  scale_y_continuous_e61(limits = c(10, 40)) +
  scale_x_continuous_e61()

save_e61(here::here("man/figures/fig-2.svg"), graph)

knitr::include_graphics(here::here("man/figures/fig-2.svg"))


```

### Density graph

```{r}
g <- ggplot(mpg, aes(cty)) + 
  geom_density(aes(fill = factor(cyl)), linetype = "blank", alpha = 0.8) +
  labs_e61(
    title = "Distribution of Car City Mileage",
    subtitle = "In miles per gallon, by number of cylinders",
    sources = "mpg dataset",
    x = NULL,
    y = "dens",
    fill = "No. cylinders"
  ) +
  mplot_label(label = c("3 cyl", "5 cyl", "6 cyl", "8 cyl"),
              x = c(25, 24, 16, 12),
              y = c(0.1, 0.35, 0.25, 0.25)) +
  theme_e61() +
  scale_y_continuous_e61(limits = c(0, 0.6, 0.2)) +
  scale_x_continuous_e61() +
  scale_fill_e61(n = 4) +
  scale_colour_e61(n = 4) +
  add_e61_logo()

save_e61(here::here("man/figures/fig-3.svg"), g)

knitr::include_graphics(here::here("man/figures/fig-3.svg"))

```

### Line graph

```{r}
df <- economics_long[economics_long$variable %in% c("psavert", "uempmed"), ]
df <- df[year(df$date) %in% c(1967:1981), ]

# plot
graph <- ggplot(df, aes(x = date)) +
  geom_line(aes(y = value, colour = variable), size = 1) +
  # Not using mplot_label just to show how plot_label works
  plot_label("Market returns", as.Date("1969-01-01"), 17, 2, 1) +
  plot_label("Unemployment", as.Date("1973-01-01"), 3, 2, 2) +
  labs_e61(
    title = "Time Series of Returns Percentage*",
    subtitle = "Drawn from long data format",
    footnotes = "Some pointless footnote",
    sources = "Economics",
    x = NULL, y = "%"
  ) + 
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  scale_y_continuous_e61(limits = c(0, 20, 5)) +
  theme_e61() +
  scale_colour_e61(2) +
  add_e61_logo()

save_e61(here::here("man/figures/fig-4.svg"), graph)

knitr::include_graphics(here::here("man/figures/fig-4.svg"))


```

### Column graph

```{r}
col_data <- head(mtcars2, 8)
col_data$names <- rownames(col_data)
col_data$index <- 1:nrow(col_data)

graph <- ggplot(col_data, aes(x = index, y = hp, fill = hp)) +
  geom_col() +
  labs_e61(
    title = "Frivolous Filled Column Chart", 
    subtitle = "For the purpose of showing off the continuous colour palette",
    sources = c("e61 Institute", "Another source"),
    x = NULL, y = "hp") +
  scale_y_continuous_e61(limits = c(0, 250)) +
  theme_e61(legend = "bottom", adj = -15) +
  scale_fill_e61(discrete = FALSE, palette = "light")

save_e61(here::here("man/figures/fig-5.svg"), graph)

knitr::include_graphics(here::here("man/figures/fig-5.svg"))

```

### Horizontal bar chart

```{r}
col_data <- head(mtcars2, 8)
col_data$names <- rownames(col_data)
col_data$index <- 1:nrow(col_data)

graph <- ggplot(col_data, aes(x = factor(index), y = hp, fill = hp)) +
  geom_col() +
  coord_flip() +
  theme_e61(legend = "bottom") +
  labs_e61(
    title = "Frivolous Filled Horizontal Bar Chart", 
    subtitle = "For the purpose of showing off the continuous colour palette",
    sources = c("e61 Institute", "Another source"),
    x = NULL, y = "hp") +
  scale_x_discrete(breaks = 1:8) +
  scale_y_continuous_e61(limits = c(0, 250)) +
  scale_fill_e61(discrete = FALSE, palette = "light") + 
  format_flip()

save_e61(here::here("man/figures/fig-6.svg"), graph)

knitr::include_graphics(here::here("man/figures/fig-6.svg"))

```

