---
title: "Introduction to theme61"
output: rmarkdown::html_vignette
description: Some example code showing how to use `theme61` with `ggplot2` graphs.
vignette: >
  %\VignetteIndexEntry{Introduction to theme61}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo = FALSE, message = FALSE, warning = FALSE}
devtools::load_all()
library(theme61)
library(ggplot2)
library(lubridate)
library(data.table)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`theme61` acts as an extension to `ggplot2`. It is made up of a custom theme and a variety of functions that change the appearance of a standard `ggplot2` graph. Thus, the correct usage of this package is to create your normal `ggplot2` graph, then add `theme61`-specific elements. This article provides some examples to help you figure out what elements you can change. Refer to these examples and the function reference page to see the full scope of this package.

## The colour palette

```{r echo = FALSE, message = FALSE, warning = FALSE}
inputs <- CJ(row = seq(1, 12), col = seq(0, 8))
inputs[, value := .I][, value := as.factor(value)]
inputs[, row := factor(row, labels = rev(c("Blue dark", "Blue light", "Teal dark", "Teal light", "Sky light", "Sky dark", "Coral dark", "Coral light", "Orange light", "Orange dark", "Maroon dark", "Maroon light")))]

prep_pal <- function(search) {
  unname(unlist(mget(apropos(search), inherits = TRUE)))
}

col_pal <- lapply(c("e61_bluedark", "e61_bluelight", "e61_tealdark", "e61_teallight", "e61_skylight", "e61_skydark", "e61_coraldark", "e61_corallight", "e61_orangelight", "e61_orangedark", "e61_maroondark", "e61_maroonlight"), prep_pal)
col_pal <- rev(col_pal)
col_pal <- unlist(col_pal)

ggplot(inputs, aes(x = col, y = row, fill = value)) +
  geom_tile() +
  coord_fixed() +
  scale_fill_manual(values = col_pal) +
  scale_x_continuous(n.breaks = 9) +
  theme_bw() +
  theme(legend.position = "none", panel.grid = element_blank(), axis.title = element_blank(), panel.border = element_blank(), axis.ticks = element_blank())
```

## Graph examples

### Scatter graph

```{r echo = TRUE, message = FALSE, warning = FALSE}

# Set up data
mtcars2 <- within(mtcars, {
  vs <- factor(vs, labels = c("V-shaped", "Straight"))
  am <- factor(am, labels = c("Automatic", "Manual"))
  cyl  <- factor(cyl)
  gear <- factor(gear)
})

# Graph
p1 <- ggplot(mtcars2) +
  geom_point(aes(x = wt, y = mpg, colour = gear)) +
  labs_e61(
    title = "Fuel Economy Declines as Weight Increases",
    subtitle = "1973-74",
    sources = "1974 Motor Trend US magazine.",
    tag = "Figure 1",
    x = "Weight (1000 lbs)",
    y = "mpg",
    colour = "Gears"
  )


p1 +
  theme_e61(legend = "bottom", legend_title = TRUE) + 
  scale_y_continuous_e61(limits = c(0, 40)) +
  scale_colour_e61(n = 3) +
  add_e61_logo() + 
  y_title_top_e61(-16) +
  square_legend_symbols()
```

### Faceted scatter graph

```{r echo = TRUE, message = FALSE, warning = FALSE}
p1 +
  facet_grid(vs ~ am) +
  theme_e61(panel_borders = TRUE, legend = "bottom", legend_title = TRUE) +
  y_title_top_e61() +
  scale_colour_e61(n = 3) +
  scale_y_continuous_e61(limits = c(10, 40)) +
  scale_x_continuous_e61()
```

### Density graph

```{r echo = TRUE, message = FALSE, warning = FALSE}
g <- ggplot(mpg, aes(cty))
g + geom_density(aes(fill = factor(cyl)), linetype = "blank", alpha = 0.8) +
  labs_e61(
    title = "City Mileage Density",
    subtitle = "By number of cylinders",
    sources = "mpg dataset",
    x = "City mileage",
    tag = "Figure 1",
    y = "density",
    fill = "No. cylinders"
  ) +
  theme_e61(legend = "bottom", legend_title = TRUE) +
  scale_y_continuous_e61(limits = c(0, 0.6)) +
  scale_x_continuous_e61() +
  y_title_top_e61(-18) +
  scale_fill_e61(n = 4) +
  scale_colour_e61(n = 4) +
  add_e61_logo()
```

### Line graph

```{r echo = TRUE, message = FALSE, warning = FALSE}
df <- economics_long[economics_long$variable %in% c("psavert", "uempmed"), ]

df <- df[lubridate::year(df$date) %in% c(1967:1981), ]

# labels and breaks for X axis text
brks <- df$date[seq(1, length(df$date), 12)]
lbls <- lubridate::year(brks)

# plot
ggplot(df, aes(x = date)) +
  geom_line(aes(y = value, colour = variable), size = 1) +
  labs_e61(
    title = "Time Series of Returns Percentage*",
    subtitle = "Drawn from long data format",
    footnotes = "Some pointless footnote",
    sources = "Economics",
    tag = "Figure 1",
    x = NULL, y = "%",
    colour = NULL
  ) + 
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  scale_y_continuous_e61(limits = c(2.5, 17.5), breaks = seq(2.5, 15, 2.5)) +
  theme_e61(legend = "bottom") +
  y_title_top_e61(-25) +
  scale_colour_e61(n = 2,
                    labels = c("Market Returns", "Unemployment")) +
  add_e61_logo() + 
  square_legend_symbols()

```

### Column graph

```{r echo = TRUE, message = FALSE, warning = FALSE}
col_data <- head(mtcars2, 8)
col_data$names <- rownames(col_data)
col_data$index <- 1:nrow(col_data)

ggplot(col_data, aes(x = index, y = hp, fill = hp)) +
  geom_col() +
  labs_e61(
    title = "Frivolous Filled Column Chart", 
    subtitle = "For the purpose of showing off the continuous colour palette",
    sources = c("e61 Institute", "Another source"),
    x = NULL, y = "hp") +
  scale_y_continuous_e61(limits = c(0, 250)) +
  theme_e61(legend = "bottom") +
  y_title_top_e61(-23, fix_left = 5) +
  scale_fill_e61(discrete = FALSE, palette = "light")
```
